---
title: "Knitting example"
author: "JR"
date: "4/18/2022"
output: github_document
editor_options: 
  chunk_output_type: console
---

# Test run of a knit

```{r setup, include=FALSE}
library(Gviz)
library(ggpubr)
library(ggplot2)
library(tidyverse)
library(GenomicRanges)
source("util/plotting_functions.R")
source("util/_setup.R")
```

# Analysis of peak features for each DBP

We've previous made a dataframe containing a number of features for each DNA binding protein and we will create some plots to examine these relationships here. KEep writting results

```{r import, message=FALSE}
# reading in num_peaks_df
num_peaks_df <- read_csv("analysis/results/num_peaks_df.csv")
```

## Histogram of peak count per DBP

```{r peak-counts}
# Distriubtion of peak numbers of all 460 DBPs
ggplot(num_peaks_df, aes(x = num_peaks)) + 
  geom_histogram(bins = 70)
summary(num_peaks_df$num_peaks)
```

The median number of peaks is nearly 16,000, but most DBPs tend to have many fewer binding events. The distribution looks like an expotential decay. Surprisingly, one DBP has nearly 150,000 peaks.

```{r peaks-vs-coverage}
# Plotting number of peaks versus total genome coverage
ggplot(num_peaks_df, aes(x = num_peaks, y = total_peak_length)) +
  geom_point() + 
  geom_smooth(method = "gam", se = TRUE, color = "black", lty = 2)+
  ylab("BP covered") +
  xlab("Number of peaks") +
  ggtitle("Peak count vs. total bases covered")
```

#### small text

```{r promoter-overlaps, message=FALSE}
# Plotting number of peaks versus peaks overlapping promoters
ggplot(num_peaks_df,
       aes(x = num_peaks, y = peaks_overlapping_promoters)) +
  xlab("Peaks per DBP") +
  ylab("Number of peaks overlapping promoters") +
  ggtitle("Relationship Between Number of DBP Peaks and Promoter Overlaps")+
  geom_point() +
  geom_abline(slope = 1, linetype="dashed") +
  geom_smooth(method = "lm", se=FALSE, formula = 'y ~ x',
              color = "#a8404c") +
  stat_regline_equation(label.x = 35000, label.y = 18000) +
  ylim(0,60100) +
  xlim(0,60100)
```

```{r genebody-overlaps, warning=FALSE, fig.width=10, fig.height=12}
# Plotting peak overlaps with genebody
ggplot(num_peaks_df,
       aes(x = num_peaks, y = peaks_overlapping_genebody)) +
  xlab("Peaks per DBP") +
  ylab("Number of peaks overlapping genes") +
  ggtitle("Relationship Between Number of DBP Peaks and Gene Body Overlaps")+
  geom_point() +
  geom_abline(slope = 1, linetype="dashed") +
  geom_smooth(method = "lm", se=F, formula = 'y ~ x',
              color = "#a8404c") +
  stat_regline_equation(label.x = 35000, label.y = 18000) +
  ylim(0,60100) +
  xlim(0,60100)
```
