---
title: "R-eading_installing_files_for_R"
author: "JR"
date: "9/27/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
library(GenomicRanges)
library(rtracklayer)
library(tidyverse)
library(ggpubr)
library(Gviz)
source("../../../util/_setup.R")
source("../../../util/intersect_functions.R")
source("../../../util/plotting_functions.R")
```

# Welcome to RStudio-viz! Let's take a little tour.

First we need to learn how to install packages. Let's find out what version of R we are using.

Simply type version in the console.

Now we will use Install.packages() for most packages. This will install any packages in CRAN repository. Let's try one.

```{R install packages}

?install.packages
install.packages("tidyverse")

```

Cool we have our first package installed.

However sometimes we ned to install from Bioconductor and need to use BiocManager::install function

```{R install BiocManager}

if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager", version = '3.10')


```

# Now try installing the packages listed above. 

Some packages like GenomicRanges are from Bioconductor and those will
need to be installed with the BiocManager::install function.

```{R install genomic ranges}

# Here is an example BiocManager::install

BiocManager::install("GenomicRanges")

# try installing all the packages listed above.

# note use either install.packages or BiocManager::install

rtracklayer
tidyverse

```


# ------------------------- R basic data analysis  -------------------------
# All the files needed for this exeRcise can be downloaded here 

```{BASH DOWNLOAD DATA NEEDED FOR LESSON IF NOT IN CLASS}

#  https://www.dropbox.com/sh/76etjxl10v2hcuc/AAAIk9a6aMhw8CvUtKlFnMn6a?dl=0

# wget (url_above) # check your directory.

```

Now let's use an example data file we will produce and how we can use R to do
some quick analyses.

We will addres the following two questions as an example:
# How many peaks does each DBP have? 
# How much of the genome do they cover?

```{r read in peaks per DBP}

# read in csv file that has number of peaks and genome coverage
peaks_per_dbp <- read_csv("../../../../../data/number_of_peaks_per_dbp.csv")

# Now we have an object in the environment that is called 'peaks_per_dbp'
head(peaks_per_dbp)

# This is a tibble or data frame object you can think of like excel sheet for now
# More on Tibble here:
# https://tibble.tidyverse.org/
```

# Using 'table' & 'summary to get information from files
```{R table and summary}

# Let's get more infomration about these peaks with table

table(peaks_per_dbp$dbp)

# Not very useful in this case because there is only one entry for each DBP

summary(peaks_per_dbp$num_peaks)
# Ok this is super helpful - we see the quartiles for each of the categories

```

# Which DBP has the most or least peaks?

```{r DBP max min peaks}

# We can take a shortcut using 'summary' to find out the smallest and largest
summary(peaks_per_dbp$num_peaks)

# we can use a greater or less than function to find the biggest one
peaks_per_dbp$num_peaks > 40000

# We see this produces the T and F for each DBP and we can see 2 DBPs (but we don't know the name :(

```

# Now we will use filter which will allow us to do much more.

```{R filter largets number of peaks}

#To solve this we can use filter and set a variable using <-
max_peaks <- dplyr::filter(peaks_per_dbp, num_peaks > 40000)

# Let's see what else we can do with filter

?filter

#Let's filter on a couple of factors
max_peaks <- dplyr::filter(peaks_per_dbp, num_peaks > 30000, total_peak_length == 133310967)

```


# Let's find the DBP(s) with the fewest peaks (we saw in summary it was 263)
```{r filter smallest number of peaks}

summary(peaks_per_dbp$num_peaks)

# we see that 263 is the min number of peaks.

min_peaks <- filter(peaks_per_dbp, num_peaks < 264)

```

# Let's find the number of peaks for a specific gene say POLR2A
```{r peaks for a specific DBP}

# let's pull out polr2a peaks 
gene_pol2 <- filter(peaks_per_dbp, dbp == "POLR2A")

```


# Now let's make a new column in the file table
```{r making a new column and populating}

# making a column of what
peaks_per_dbp$newcol <- "What?"

```

# Now let's assign a number to each DBP in another new column
```{R new column populating with a function}

peaks_per_dbp$id <- 1:nrow(peaks_per_dbp)

```

# Indexing
```{r indexing rows 1:5 and column 1}

peaks_per_dbp[1:5, 1]

# IMPORTANT : R is always indexes by rows, then columns (comma seperates)
# So above we see take rows one through five and column 1 information.

# More on R indexing https://stats.idre.ucla.edu/r/modules/subsetting-data/

```

# Let's remove a column using indexing
```{R remove a column using indexing}

# let's remove the "what" column we just made above
peaks_per_dbp <- peaks_per_dbp[,-4]

```


# Now let's make a plot of the data. First a histogram of number of peaks for each DBP
```{r}
?hist

hist(peaks_per_dbp$num_peaks, breaks = 40) 

# saving the figure (we will do this over and over again :)
# we will learn ggplot later but we use PDF and turn of null device (plot window) 

pdf("figures/hist_of_peaks_per_dbp.pdf")
hist(peaks_per_dbp$num_peaks, breaks = 40) 
dev.off()

# note always check working directory before saving.
```

# Now let's plot the number of peaks versus peak width.
```{r number of peaks versus genome coverage}

# we will use base R to plot the num_peaks versus genome coverage

pdf("figures/num_peaks_vs_coverage.pdf")

plot(x = peaks_per_dbp$num_peaks, y = peaks_per_dbp$total_peak_length)

dev.off()

```

# Now let's dig into the data file to subset it into the genes we are interested in.
```{r selcting a list of genes from data frame}

# First let's make a list of genes we want. 
# To make a vector of characters we use the c() function

genes <- c("POLR2A","POLR2B", "SUPT5H")

# Now lets make it a searchable index with paste0 (?paste0)
pattern_to_search <- paste0(genes, collapse = "|")

```

# It's worth noting that things can be accomplished in many different ways in R.
# Here we'll get to the same subset of row entries two different ways. 
```{r}
#first let's use grepl
# ?grepl

grepl(pattern_to_search, peaks_per_dbp$dbp)

# table will give us a way to test if we retrieve the three DBPs

table(grepl(pattern_to_search, peaks_per_dbp$dbp))

# Now let's get all the information for these genes by indexing peaks_per_dbp

# So df[,] = all rows/cols, df[1,] = first row all columns etc.

ppdbp_subset <- peaks_per_dbp[grepl(pattern_to_search, peaks_per_dbp$dbp),]

# note we only took the rows for the genes of interest. Then all the cols with (,)
# R syntax is rows,columns.
# df[,] = all rows/cols
# df[1,] = first row all columns
# df[,1] = all rows and column 1

```

# Now let's do the same thing with filter in tidyverse syntax ' %in% ' 
```{R using %in%}

# %in% will find all the data "in" the list being searched "genes"

peaks_per_dbp$dbp %in% genes

# Now we will set a variable as the output using filter and then the code we just ran

ppdbp_subset <- dplyr::filter(peaks_per_dbp, dbp %in% genes)

```

# MUTATE !!


# Now let's run functions on the data in a given column with MUTATE (my favorite function)
# head(peaks_per_dbp) -- we can see that the total length is quite long, let's make it in Millions
```{r mutate}

#let's take a look at mutate
?mutate

# Now let's use mutate to divide the total coverage by 1M for every value
peaks_per_dbp <- peaks_per_dbp %>%
  dplyr::mutate(tpl_m = total_peak_length / 1000000)

# syntax is mutate(new_col_name = function(x) in this case divide :)

# instead of making a new col we can set mutate output to old col name

peaks_per_dbp <- peaks_per_dbp %>% 
  mutate(total_peak_length = total_peak_length / 1000000) 

# Now let's plot it
hist(peaks_per_dbp$total_peak_length)

```

# Now let's explore the power of mutate syntax with a function.
# We will try it with a more complex function using ifelse logic

```{r mutate if else statement}

peaks_per_dbp <- peaks_per_dbp %>% 
  mutate(genome_coverage_size = ifelse(total_peak_length < 1, "small",
                          ifelse(total_peak_length < 30, "medium", "large")))

# that is really handy for changing stuff around in a data frame -- 
# note that if else moves forward for greater that 30 and if true then returns large
# In other words if else you only need to conditions to get three values ...
```


*********************
EXCERCISE
*********************

Use what we learned above to plot a historgram of all the proteins that are in the medium sized category for total genome coverage.

Make a table with just the medium coverage DBPs
