---
title: "our_data_in_ranges"
author: "JR"
date: "11/2/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(GenomicRanges)
source("../util/intersect_functions.R")
source("../util/plotting_functions.R")
source("../util/_setup.R")
library(Gviz)
library(ggplot2)
```


Alright time to start digging into the data for new results !!
We have made a consensus peak file for each DBP and 
Moreover, we have a data frame of a bunch of peak overlap features for each DBP.

First let's load in the consensu peak files (this is something we always have to do)
```{r}

fl <- list.files("../data/consensus_peaks", full.names=TRUE)
consensus_peaks <- lapply(fl, rtracklayer::import, format = "bed")


# Let's take a look since this is an important first step.
# Ok we should probably add some names -- another common task.
# we can get name from @elementdata$listdata$name

consensus_peaks[[40]]$name %>%
  head()

# cool now we can strsplit this to get the name on the left of _

names(consensus_peaks) <- sapply(consensus_peaks, function(x){
  unlist(strsplit(x$name, "_"))[[1]]
  
})

# Now we have a named list of consensus peaks in the environment -- let's take a look
```


Let's load in our num_peaks_df and decide if we want to filter?
```{r}

# loading in our previous analyses:

num_peaks_df <- read.csv("/Shares/rinn_class/data/CLASS_2022/class_exeRcises/analysis/13_ChIP_peak_features/num_peaks_df.csv")
  
# if we want to filter to those DBPs with at least 500 peaks:

num_peaks_threshold <- 500
dbps_with_sufficient_peaks <- num_peaks_df$num_peaks > num_peaks_threshold

# let's see how many we loose doing this:
summary(dbps_with_sufficient_peaks)

# Note that the enviroment var is still 483 long as it is boolean.
# Let's filter our consensus peaks GRanges list

filtered_consensus_peaks <- consensus_peaks[sapply(consensus_peaks, length) > num_peaks_threshold]

# Could also filter like this:

filtered_consensus_peaks <- consensus_peaks[dbps_with_sufficient_peaks]

# we can record the result of what percentile this cutoff is at :
threshold_percentile <- round(ecdf(num_peaks_df$num_peaks)(num_peaks_threshold)*100,1)

```

Ok enough set up let's dig in with making some plots!

```{r}

# First let's look at a histogram of peak#/DBP

 ggplot(num_peaks_df, aes(x = num_peaks)) + 
  geom_histogram(bins = 70)

# We can add a bunch of example geom_() functions

ggplot(num_peaks_df, aes(x = num_peaks)) + 
  geom_histogram(bins = 70) +

  geom_vline(xintercept = num_peaks_threshold, lty = 2,
             size = 1.1, color = "#a8404c") +
  annotate(geom = "text", x = num_peaks_threshold + 3e3, y = 35, 
           color = "#a8404c", label = paste0("t=", num_peaks_threshold)) +
    annotate(geom = "text", x = num_peaks_threshold + 2.2e3, y = 32, 
           color = "#a8404c", label = paste0(round(threshold_percentile,0), "%")) +
  xlab("Number of consensus peaks") +
  ylab("Count") +
  ggtitle("Distribution of number of consensus peaks")
  
# Now let's save this figure as a result using ggsave.
# ** important to make 'figures' dir in analysis 14_2
  

ggsave("/Shares/rinn_class/data/CLASS_2022/class_exeRcises/analysis/14_2_plotting_our_data/figures/consensus_peaks_histogram.pdf")

# we will save figures in: /analysis/X_RMD/figures so we have a dir
# for each RMD output that has figures corresponding to that code.
```




```{r}

# Let's plot our result of num_peaks versus genome coverage.

ggplot(num_peaks_df, aes(x = num_peaks, y = total_peak_length, label = dbp)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "black", lty = 2,
              formula = 'y ~ x') +
  geom_text() +
  ylab("BP covered") +
  xlab("Number of peaks") +
  ggtitle("Peak count vs. total bases covered")

ggsave("/Shares/rinn_class/data/CLASS_2022/class_exeRcises/analysis/14_2_plotting_our_data/figures/peak_num_vs_coverage.pdf")

# Feel free to copy code from 14_plotting_R_part_I.Rmd for other versions of this figure.

```


Let's did in a bit deeper and see how coverage on promoters.
```{r}
# now let's plot x as num_peaks and y num overlapping promoters.

ggplot(num_peaks_df,
       aes(x = num_peaks, y = peaks_overlapping_promoters)) +
  

  xlab("Peaks per DBP") +
  ylab("Number of peaks overlapping promoters") +
  ggtitle("Relationship Between Number of DBP Peaks and Promoter Overlaps")+
  geom_point() +
  geom_abline(slope = 1, linetype="dashed") +
  geom_smooth(method = "lm", se=F, formula = 'y ~ x',
              color = "#a8404c") +
  stat_regline_equation(label.x = 35000, label.y = 18000) +
  ylim(0,60100) +
  xlim(0,60100)

ggsave("/Shares/rinn_class/data/CLASS_2022/class_exeRcises/analysis/14_2_plotting_our_data/figures/3_peak_num_vs_promoter_coverage.pdf")
```


Coverage on gene bodies
```{r}


ggplot(num_peaks_df,
       aes(x = num_peaks, y = peaks_overlapping_genebody)) +
  

  xlab("Peaks per DBP") +
  ylab("Number of peaks overlapping genes") +
  ggtitle("Relationship Between Number of DBP Peaks and Gene Body Overlaps")+
  geom_point() +
  geom_abline(slope = 1, linetype="dashed") +
  geom_smooth(method = "lm", se=F, formula = 'y ~ x',
              color = "#a8404c") +
  stat_regline_equation(label.x = 35000, label.y = 18000) +
  ylim(0,60100) +
  xlim(0,60100)

# Interesting result !! Gene bodies explain amost all the places of binding in 
# the genome! Where as promoters had a non linear asymptope after a certain number 
# of peaks. VERY COOL and surprising. There are billions of other possible binding sites!

ggsave("/Shares/rinn_class/data/CLASS_2022/class_exeRcises/analysis/14_2_plotting_our_data/figures/4_peak_num_vs_gene_body_coverage.pdf")
```

Ok so we have started generating results in this handy num_peaks_df. 
However, the peak_occurence_matrix is also very handy for analyses.
Let's load it in and do some row and col sums.

```{r}

peak_occurence_df <-



```




```{r Density plot of number of DBPs per promoter}
# Since we are making a density plot we only need the X axis.
# this is the number_of_dbp -- which we added to the dataframe above using
# colsums function across the promoter peak occurence matrix. So it essentially
# returns the number of times a DBP overlaps a promoter (we didn't include genebody)
g <- ggplot(peak_occurence_df, aes(x = number_of_dbp))
#AES layer will know to make a histogram since there is only one set of values for x.
g + geom_density(alpha = 0.2, color = "#424242", fill = "#424242") +
  theme_paperwhite() +
  xlab(expression("Number of DBPs")) +
  ylab(expression("Density")) +
  ggtitle("Promoter binding events",
          subtitle = "mRNA and lncRNA genes") 
# Wow what a result! Last year we only saw up to 111 binding events at a promoter
# this is kind of validated where the first peak in the distribution falls at 100
# binding events -- yet there is a second distribution of regions of 300 or more 
# binding events -- wonder what those are :) ?
```


Lets find how many promoters don't have any DBPs bound

```{r determining if a promoter is never bound}
unbound_promoters <- peak_occurence_df %>% filter(peak_occurence_df$number_of_dbp < 1)
nrow(unbound_promoters)
# here is just a simple index and filter of the index to have at least 1 dbp bound.
write_csv(unbound_promoters, "results/unbound_lncrna_mrna_promoters.csv")
```


Do any DBPs never bind a promoter?

```{r determing how many DBPs have no overlaps with promoters}
# Determining how many DBPs don't bind to any promoters.
dbp_promoter_ovl <- get_overlapping_peaks(lncrna_mrna_promoters, filtered_consensus_peaks)
num_ovl <- sapply(dbp_promoter_ovl, length)
min(num_ovl)
summary(num_ovl)
table(num_ovl)
# There are no DBPs that have no promoter overlaps. The minimum number of promoter
# overlaps is 54
```


```{r comparing lncRNAs and mRNA promoter overlaps}
# Class exercise, try and make sense of what was done here based on the above
num_peaks_dfl <- num_peaks_df %>%
  dplyr::select(-peaks_overlapping_promoters) %>%
  pivot_longer(cols = peaks_overlapping_lncrna_promoters:peaks_overlapping_mrna_promoters,
               names_to = "gene_type",
               values_to = "peaks_overlapping_promoters") %>%
  mutate(gene_type = gsub("peaks_overlapping_", "", gene_type))
ggplot(num_peaks_dfl, aes(x = num_peaks, y = peaks_overlapping_promoters, 
                         col = gene_type)) +
         geom_point() +
         geom_abline(slope = 1, linetype="dashed") +
  geom_smooth(method = "lm", se = FALSE, formula = "y ~ x") +
  stat_regline_equation() +
  scale_color_manual(values = c("#a8404c", "#424242"))+
  xlab("Peaks per DBP") +
  ylab("Peaks Overlapping Promoters") +
  ggtitle("Number of DBP Peaks and Promoter Overlaps")
ggsave("figures/peaks_overlaps_relationship_by_gene_type.png", height = 5, width = 8)
ggsave("figures/peaks_overlaps_relationship_by_gene_type.pdf", height = 5, width = 8)
``` 
